Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals

### 2022-06-01 draft_BLC/blc_reviewer_response.txt

Manuscript BLC-21-0133
Response to Reviewers

Dear Editors,

Thank you for the opportunity to submit a revised draft of our manuscript “Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals” for publication in Bilingualism: Language and Cognition. We appreciate the time and effort from your end and that of the reviewers in providing feedback on our manuscript, and in particular the thoughtful comments for improvements to the work. We have incorporated the suggestions made by the reviewers. Please see below, for a point-by-point response to the reviewers’ comments and concerns. All page numbers refer to the revised manuscript file uploaded in the portal.

Sincerely,
Auromita

------------------------------------------------

Reviewers' comments to authors:

Reviewer: 1

I will focus my comments here on a few things I believe should be improved. Let me say first that I am quite impressed by the quality of the writing and the insight demonstrated. This is a very fine project and it has many positive things going on for it. There are a number of issues the authors would do well to address, though.

1. The sample size is just too small. The study recruited only 10 participants. This is not much less than comparable studies, I know. The standards in our field, however, are quickly changing. This study sets out to compare the sizes of the effects of two experimental treatments or conditions. Effect sizes are likely to be very small in the population, if they exist at all. This is designed as a confirmatory study, rather than an exploratory study. It is one thing to explore an issue for the first or second time, and another to confirm a specific effect others believe they have found while exploring the possibility of its existence. A confirmatory experiment needs a power analyses, or at the very least the sample size must be explicitly justified. I suspect the sample is too small to detect any differences between the two treatments (picture naming vs. code-switching), and the differences between the unilingual and the bilingual trials is apparently found to be very small. The likelihood of detecting a significant difference increases with a more powerful experimental design; and if no differences are detected one may propose that none exist in the population. And, for significant differences, a more accurate estimate can be found in more powerful designs--those with larger samples. I'd like to encourage the authors to consider the possibility to return to the field and to continue to collect data, if at all feasible. Otherwise, at the very least, acknowledge the limitations openly and justify your sample size. Changes are this won't be enough, but only the editorial team can decide whether they continue to accept low-power studies or not.

Authors' response (AR): We agree that the sample size is smaller than ideal. We were constrained by resources and participant availability given a stringent inclusion criteria (discussed in response point 2).This was in order to minimize random variability in the sample, given that the population we are working with is highly variable in terms of linguistic experience. Unfortunately, it is not possible to collect more data. Our university has still not fully reopened after the pandemic, and we therefore don't have access to a lab at present. We plan to expand this study by collecting data online in the near future, but that is beyond the scope of the present study. 
We have performed a power analysis and included it in our manuscript (section 3.4, page 35). This shows that the study does have enough power to detect the effects of interest (phonetic transfer: vowel*context interaction, and paradigm differences: vowel*context*task interaction in F1). We acknowledge that the power for detecting a paradigm effect in F2 is low, given the effect sizes in the present data. We have acknowledged the sample size limitation within the text (page 36). However, we would like to note a couple of points:

i) There is a body of literature that debates the usefulness of post-hoc power analyses in interpreting the findings of a completed study where statistical analyses have already been performed:

Lakens, D. (2021, January 4). Sample Size Justification. https://doi.org/10.1525/collabra.33267
Dziak, J. J., Dierker, L. C., & Abar, B. (2020). The interpretation of statistical power after the data have been gathered. Current Psychology, 39(3), 870-877.
Lenth, R. V. (2007). Post hoc power: tables and commentary. Iowa City: Department of Statistics and Actuarial Science, University of Iowa, 1-13.
Gelman, A. (2019). Don’t calculate post-hoc power using observed estimate of effect size. Annals of surgery, 269(1), e9-e10.

This is especially important in situations where norms around expected effect sizes are yet to be established in the field, and thus cannot be readily motivated on theoretical grounds, making it necessary to rely on observed effect sizes from the data.

ii) We still do not know enough about power calculations in complex mixed effect models and effect sizes for cross-language transfer phenomena to make designing and interpreting such analyses straightforward. We have discussed this in some detail in the manuscript (page 36). Although we have done our best to conduct a power analysis with minimum bias, we are open to any methodological suggestions the reviewers might have.

iii) In light of these considerations, we want to draw your attention to the scarcity of literature on language behaviors of native multilingual populations. This makes it doubly difficult to find effect size estimates from comparable populations, and more generally to rely on existing norms/expectations while making methodological decisions. In large part, this is due to resource inequality, leading to a system that propagates more research on already well-studied languages and populations. The norms for experimental research on bilingualism have been established in a Western context. In a setting where linguistic variability at the population level is much greater than canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. For example, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7%)  in the final study given our inclusion criteria. Note that these criteria (discussed further later) establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most contemporary transfer studies. This example demonstrates the degree of linguistic variability in the present population. In such a scenario, is it scientifically justified to apply blanket uniform standards for participant numbers across populations?  
We completely agree that power considerations are important, and will strive for more well-powered designs in future studies. But we also wish to point out that the lack of diversity in existing bilingual data is an equally large impediment to progress in our field. We believe that the norms that propagate this urgently need to change. Given the methodological uncertainties discussed above, the interpretation of this power analysis is tentative at best (c.f. discussion on page 35). On the other hand, existing results from transfer studies clearly demonstrate the need for data from a wider range of populations and language pairs (c.f. discussion in section 1, Introduction). While we respect whatever decision the editorial team makes in this regard, we write this to draw your attention to an alternative perspective on this issue.

2. We need to know a lot more about the participants. Participant information takes a single paragraph. Readers are not necessarily familiar with this particular bilingual population. What's more, there are no indicators of bilingualism or English or Bengali proficiency. The authors would do well to collect information from the participants utilizing one of the available bilingualism surveys, such as the BLP  or the LEAP-Q. Also, linguistic proficiency should be assessed somehow; perhaps a vocabulary size measure or a grammar knowledge measure. Or at lest a self-assessed proficiency questionnaire. This is of course optional, particularly in native bilingual populations. In this case, though, the reader needs to learn a lot more about the society and the bilingual conditions of both the population and the specific sample: language use patterns, ages and contexts of acquisition, literacy, origin, ages, education levels and languages of education, etc. Finally, consider the possibility of reporting the inclusion and exclusion criteria openly and explicitly.

AR: Thank you for pointing this out. As per reviewer suggestions, we have discussed the linguistic situation of the population in greater detail (section 2.1, page 20) . We had collected some language background information from participants prior to the study, which was used to screen participants. We have now collected additional information about literacy, acquisition, usage habits and attitudes (in the form of a collated survey of the BLP + LEAP-Q + Bilingual Switching Questionnaire (BSWQ)) from the participants of the study (we were able to reach 9 of the 10 original participants). We have discussed these responses in the Participants section. This language background survey includes self-ratings for proficiency, which have been shown in existing literature to correlate well with actual performance. Moreover, educational history is a reliable predictor of proficiency for this population, as detailed in the manuscript. We have reported our inclusion criteria, and discussed the motivation behind it.

3. The Discussion section is brief and shallow. I am impressed by the quality of the writing in the Introduction, but the Discussion and Conclusions are no match for it. The Discussion needs to be much longer, much deeper, it needs to engage with theory and contextual findings explicitly and in depth, it needs to interpret the specific findings and explain their implications. I suggest that the "limitations" paragraph be moved out of the last position in the article. It could perhaps be placed in the Discussion section, but not in the Conclusions section. The Conclusions section needs to focus on the positive findings: what have we learned here? What can we say after having conducted this study? Etc. Finally, regarding the Discussion section, there should be a separate subsection entitled "Summary of Findings" (of course, whether there is such a separate section or not is a personal matter, but my point is that there need to be a few paragraphs in which the authors summarize the findings (perhaps by providing standardized effect sizes) in plain English while refraining from interpreting or explaining the data. There needs to be a subsection in which the data are summarized so that the reader can remember the findings before engaging in a discussion of their implications.

AR: We agree with this feedback. Accordingly, we have now rewritten the Discussion section to be longer and more detailed, with an overall summary of findings at the beginning (page 37), and separate subsections for each of our original hypotheses. We have added a separate subsection titled “Model output summary” in the Results section where we report the fitted model, model output, and effect sizes without interpretation. This is followed by a “Model interpretation” subsection which explains the effects in plain language without theoretical interpretation – i.e.  what the numbers mean in terms of acoustics (e.g. the negative lope of the vowel*context interaction indicates that \nt is lowered to a greater extent than \ae in the mixed condition). In the Discussion section, we have interpreted the finding, discussed it in light of our original research questions and hypotheses, and discussed theoretical implications.

Except for my suggestion to keep collecting data to obtain a larger sample (as well as collecting more information from the participants, in the form of some bilingualism survey), I believe my other suggestions are relatively minor. I must say that I was quite impressed while reading the manuscript until (1) I saw the Participants section (small sample size and sparse information about the linguistic histories and experiences of the participants), and (2) I read the Discussion (brief, shallow, avoiding a deep discussion of the implications of the study and lacking a summary of findings devoid of interpretation before moving on to their interpretation). I sincerely hope the authors are willing to consider some of my comments. I believe this is, overall, a fine study.

AR: Thank you for these suggestions and feedback! We hope we have been able to address the major concerns.

------------------------------------------------------------------------------------------
Reviewer: 2

This study set out “to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant’s multilingual baseline production of L2.” The major claim (page 24) is that “Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.”
I am concerned about whether the results actually support this claim. As the ms. makes very clear, the data on baseline production of these vowels in IE was taken from a different study of five female speakers.

AR: As outlined in the Methodology section (page 19), our study had a within-participant design: we elicited IE speech from a group of bilingual speakers in two conditions-- unilingual (IE), and mixed (switching from Bengali to IE). Our statistical analyses compared productions across these two conditions. This means that participants' unilingual IE productions were compared to their own mixed productions, and were found to differ in vowel quality. Thus, the main claim that “Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.” is certainly supported by the results.
Based on previous literature, we expect that such shifts in vowel categories are not random, but rather targeted with respect to L1 categories. Although our primary research question is about whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. Towards this, we used existing data to estimate average formant values of the relevant Bengali vowel categories. Initially, this data was taken from Dutta (2021). However, in light of concerns about the sample size and gender makeup of that data, we have now used a larger dataset consisting of both male and female speakers from the SHRUTI corpus[i] (c.f. response to comment 1 of reviewer 3). The distribution of vowels was found to be the same in both datasets. Page 16 of the revised manuscript contains a detailed discussion of how this data was extracted and used, including speaker information from the corpus. Note that we have clearly mentioned that this baseline Bengali data represents average formant values, and is only used as a heuristic for discussion. We have not included this data in our statistical analyses. 
[i] Das, B., Mandal, S., & Mitra, P. (2011, October). Bengali speech corpus for continuous auutomatic speech recognition system. In 2011 International conference on speech database and assessments (Oriental COCOSDA) (pp. 51-55). IEEE.

The participants in this study were ten male speakers. While we are told that the speakers in this study “were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW...skills in L1 and L2 across participants), we are not given any information on the criteria used in this selection, and no information beyond gender is provided for the baseline group.

AR: We have taken this feedback into account, and reported the inclusion criteria for the study, and the motivations for using those specific metrics for participant selection (section 2.1, page 20; c.f. response to comment 2 of reviewer 1). More information about the baseline Bengali data, including speaker information, was added as a separate subsection on page 16 (c.f. response to comment 1 of reviewer 3).
(We would like to note that the participants in our study were 5 male and 5 female speakers, rather than 10 male speakers, as detailed in section 2.1 (page 20).)

The ms. seems to assume that the 5 female speakers in the baseline study can be taken as representative of all L1 IE speakers, but we have no evidence for that.

AR: We reiterate that the baseline data was for L1 Bengali vowels. However, in light of concerns about the small sample size, we have now used a larger dataset of Bengali speech extracted from the SHRUTI corpus, which contains both male and female speakers, and is expected to be representative of vowel targets of educated Western Bengali speakers from India (c.f. response to comment 1 of reviewer 3).

A second claim (p. 25) is that “As expected, we find a greater degree of shift in [ʌ], possibly owing to its position in a less dense part of the IE vowel space.” This is a very interesting claim but it would be much more convincing if we were provided explicit information on the IE vowel space, in the form of a chart of all the vowels in this language. We also need to know whether acoustic differences in addition to F1 and F2 (such as duration) play a role in distinguishing vowels.

AR: Thanks for this suggestion! We have accordingly added a vowel chart for IE (page 60), and discussed the two categories of interest, [ʌ] and [æ], in greater detail in section 4.2, page 44. As for other acoustic features, while the IE vowel system uses phonological length contrast, the two categories studied here are not part of a length-contrasted minimal pair (they do not have a long or short counterpart). While [ʌ] and [æ] differ in their canonical durations, with [æ] being a longer vowel[i], we are not aware of any perception data from L1 Bengali IE speakers that could illuminate the role of duration as a perceptual cue to these categories.  
[i]Maxwell, O., & Fletcher, J. (2009). Acoustic and durational properties of Indian English vowels. World Englishes, 28(1), 52-69.

Finally, at a number of points the ms. makes reference to the lack of studies of cross-language influence on vowels (as opposed to VOT). I’m not sure this is a fair characterization. While studies of the effect of language switching on vowels may be lacking, there are certainly a number of studies investigating phonetic drift in vowels, for example:
Grace E. Oh, Susan Guion-Anderson, Katsura Aoyama, James E. Flege, Reiko Akahane-Yamada, Tsuneo Yamada. 2011. A one-year longitudinal study of English and Japanese vowel production by Japanese adults and children in an English-speaking setting. Journal of Phonetics 39: 156-167.
Natalia Kartushina and Clara D. Martin. 2019. Third-language learning affects bilinguals’ production in both their native languages: A longitudinal study of dynamic changes in L1, L2 and L3 vowel production. Journal of Phonetics 77: 100920.
Benjamin Lang and Lisa Davidson. 2019. Effects of Exposure and Vowel Space Distribution on Phonetic Drift: Evidence from American English Learners of French. Language and Speech 62:30-60.

AR: Thank you for pointing us to these references. Our discussions about the lack of studies focusing on vowels are very specifically about short-term, transient phonetic interaction, such as is triggered by language mixing or immediate language context. Such studies demonstrate local cross-language influence during speech processing, as opposed to changes to the mental representations of sound categories over longer periods. We distinguish between long- and short-term transfer early in the text (Introduction, page  3), and focus most of our literature review on the latter, unless explicitly specified otherwise. In light of the reviewer's concern, we have rephrased the text on page 10 ("all the studies of mixed-language production discussed above measure VOT, which is a temporal feature.") to indicate this more clearly.

Specific comments
page 6: I found the sentence in (ii) confusing.
AR: Original sentence: "Given that examples of an IE phonology without L1 'influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition."
As discussed in the preceding paragraph (page 6.), for the majority of this population, English is learned and used as an L2 to communicate with other L2 speakers. There are very limited L1 IE speakers. Thus, what is collectively understood as "Indian English" is not a monolithic entity, as it is always colored by a speaker's other languages. Examining cross-language interaction in this scenario amounts to observing an increase in the amount of such L1 "accentedness" in IE speech.

page 6: ‘monopthongs’ should be ‘monophthongs’
page 9: First sentence: “bilinguals speakers’...”
page 14: “individual sounds pairs”
AR: Thanks for pointing these out, we have edited these lines in the manuscript.

page 9: “The research discussed above concerns situations where the bilingual speaker is
operating in any one of their languages.” It seems odd to refer to studies investigating the first
five weeks of learning a language as involving bilinguals.
AR: We have rephrased this line on page 10: "The research discussed above concerns situations where a speaker is operating in any one of their languages"

page 13: It wasn’t clear to me how the discussion of the Bengali laryngeal contrast was relevant
here.
AR: We discuss the four-way laryngeal contrast to motivate our use of vowels as the feature of interest in this study, rather than VOT. The large body of literature on phonetic transfer in VOT is largely based around languages where VOT is the primary cue for voicing contrast (voiced vs voiceless). Thus, exploiting canonical VOT differences between languages allows us to ascertain cross-language influence. In Bengali, the voicing contrast is four-way (voiced vs voiceless, aspirated/breathy vs unaspirated), and VOT is not the primary cue for this contrast. If we were to examine changes to L2 IE VOT as a result of L1 Bengali influence, it would be difficult to interpret the observations in a straightforward manner, owing to the additional phonological differences between the languages.

------------------------------------------------------------------------------------------------------------------
Reviewer: 3

Comments to the Author
The paper reports two tasks – cued-picture naming and code switching – investigating cross-language phonetic transfer in bilingual (Bengali-English) speakers. The study focused on the F1 and F2 of two vowels of the L2 (Indian English) language. Both vowels were subject to changes, with a shift toward the L1 norms. The same changes occurred independently of the task. The study thus reports first evidence of transfer effects on L2 vowels.

I've found the study interesting, and the findings having a good potential for significantly contributing to models of bilingual language control. There are some relatively minor issues (mostly methodological) that I think deserve attention and need to be addressed.

1. As baseline data, the authors use Dutta's (2021) dataset, which is composed of production data from 5 female speakers. In the present study, however, participants are 5 females and 5 males. Because of the effects of gender on vowel realization, do the authors think that this difference between the two groups (Dutta vs the present one) might have contributed to the effects reported in the present study?

### 2022-06-01 BLC_manuscript_1/BLC_AM-ID_manuscript.tex
BLC revision final

% !TeX document-id = {6d544fe2-4255-4108-8022-d809ff1d2c53}
% TeX program = xelatexmk %doesn't work, no xelatexmk option under 'build'. Changed default engine to xelatex (options-->configure-->build-->default build-->xelatex)
% see http://info.semprag.org/basics for a full description of this template
\documentclass[charis,linguex]{glossa}
% possible options: 
% [times] for Times font (default if no option is chosen)
% [cm] for Computer Modern font 
% [lucida] for Lucida font (not freely available)
% [brill] open type font, freely downloadable for non-commercial use from http://www.brill.com/about/brill-fonts; requires xetex
% [charis] for CharisSIL font, freely downloadable from http://software.sil.org/charis/
% for the Brill an CharisSIL fonts, you have to use the XeLatex typesetting engine (not pdfLatex)
% for headings, tables, captions, etc., Fira Sans is used: https://www.fontsquirrel.com/fonts/fira-sans
% [biblatex] for using biblatex (the default is natbib, do not load the natbib package in this file, it is loaded automatically via the document class glossa.cls)
% [linguex] loads the linguex example package
% !! a note on the use of linguex: in glossed examples, the third line of the example (the translation) needs to be prefixed with \glt. This is to allow a first line with the name of the language and the source of the example. See example (2) in the text for an illustration.
% !! a note on the use of bibtex: for PhD dissertations to typeset correctly in the references list, the Address field needs to contain the city (for US cities in the format "Santa Cruz, CA")

%\addbibresource{sample.bib} 
% the above line is for use with biblatex
% replace this by the name of your bib-file (extension .bib is required)
% comment out if you use natbib/bibtex

%for natbib: replace \cite with \citep (parenthetical); \citea with \cite (in-text)
\usepackage{hyperref}
\let\B\relax %to resolve a conflict in the definition of these commands between xyling and xunicode (the latter called by fontspec, called by charis)
\let\T\relax
\usepackage{xyling} %for trees; the use of xyling with the CharisSIL font produces poor results in the branches. This problem does not arise with the packages qtree or forest.
\usepackage[linguistics]{forest} %for nice trees!

@@

%\graphicspath{{images/}}

% \pdf* commands provide metadata for the PDF output. ASCII characters only!
%\pdfauthor{Auromita Mitra, Indranil Dutta}
%\pdftitle{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
\pdfkeywords{phonetic transfer, vowel formants, Bengali, English}

\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
% Optional short title inside square brackets, for the running headers.

@@

\begin{document}

\sffamily
\maketitle

Word count: 9263

\begin{abstract}
This study investigates the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and English, to better understand the nature of cross-language phonetic interaction. Existing research has largely focused on a small set of phonologically related languages and used voice onset time (VOT) to measure phonetic transfer. Results indicate that such interaction is mediated by both language-specific features and linguistic experience, highlighting the need for diversification in the range of languages and populations studied. We compare productions from a yet-unstudied population in two conditions-- unilingual L2 (English) and mixed (Bengali and English). The spectral properties of two L2 vowel categories— \nt{\ae} (common to L1 and L2) and \nt{2} (absent in L1) are examined as sites for transfer. To understand how this unfolds dynamically, we measure F1 and F2 at five points over the course of articulating the vowel. We find dynamic and asymmetrical changes in both vowels during mixed production, with \nt{2} shifting towards a related L1 category \nt{a:}. We interpret this as a temporary increase in cross-language phonetic influence during mixed-language use, leading to a shift towards L1 norms. Mixed language is elicited in two switching paradigms: cued picture-naming and code-switching, to assess if differences between the paradigms independently influence the outcome of phonetic interaction. Results reveal no difference in transfer patterns across the two paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the postulated role of connected speech in introducing discursive factors to transfer studies. 
\end{abstract}

\begin{keywords}
 phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali 
\end{keywords}

\rmfamily

\section{Phonetic transfer}

Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}. However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language ``systems" are cognitively represented and processed. Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
%show cross-language-- rephrase

%keep just the types here, and put the next 2 lines in the causes&effects section?
Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production}; 
(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}. %These compare productions during mixed-language use to participants' own productions while using a single language. The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness. 
The present study is concerned with the latter. In production, this kind of influence has been variously termed \textit{transfer, drift, accommodation,} and \textit{interference}-- we use these interchangeably here to indicate any interaction between two sets of phonetic norms.\\

Existing research on the phonetic effects of mixed-language production has largely focused on a limited set of phonologically related languages. A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer. However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants. This emphasizes the importance of considering data from a wider variety of
%elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer. 'Types of features'--separate line?- Brief mention - discursive factors in VOT. 
populations, language pairs,  and phonetic features in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
%line--rephrase
behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing. However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language. \\

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.\\

The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems. \ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs the local act of switching between languages. Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter. The design of the present study is motivated on the basis of this discussion. A logical consequence of this model is that such phonetic interaction must be highly localized. \ref{duration} elaborates on this by reviewing literature on the duration and ``reversiblity" of transfer effects. A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies. In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that mediate such phonetic interaction (or fail to do so). We argue that these results point to an urgent need for expanding the range of language pairs and sound categories examined in order to make meaningful generalizations about the nature of short-term phonetic transfer. Following from this, \ref{asymmetry between sounds} discusses reported asymmetries between different sound categories of a single language, and develops the hypotheses of the present study on the basis of these. In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching. \ref{questions_and_hypotheses} summarizes the research questions and hypotheses.

@@

10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks. All the participants speak other languages in addition to Bengali and English. At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population. Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW skills in L1 and L2 across participants, and were compensated for their time.

>>>>

% !TeX document-id = {6d544fe2-4255-4108-8022-d809ff1d2c53}
% TeX program = xelatexmk %doesn't work, no xelatexmk option under 'build'. Changed default engine to xelatex (options-->configure-->build-->default build-->xelatex)
% see http://info.semprag.org/basics for a full description of this template
%\documentclass[times,linguex]{glossa}
\documentclass[12 pt]{article}
\usepackage[letterpaper]{geometry}
\usepackage{times}
\geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}

\usepackage[hidelinks]{hyperref}
%\renewcommand{\sectionautorefname}{\S}
\usepackage{xcolor}
\hypersetup{
	colorlinks,
	linkcolor={blue!50!black},
	citecolor={blue!50!black},
	urlcolor={blue!80!black}
}

\usepackage{graphicx} %for images
\graphicspath{{images/}}
\usepackage{caption}
\captionsetup[table]{skip=0pt,singlelinecheck=off}
\captionsetup[figure]{skip=0pt,singlelinecheck=off}

\usepackage{booktabs}
\usepackage{enumitem} %customize list numberings

\usepackage[natbibapa]{apacite}
\usepackage{natbib}

@@

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[R]{\thepage}

\usepackage{setspace}
\doublespacing

\usepackage{comment}

\newlength\mystoreparindent
\newenvironment{myparindent}[1]{%
	\setlength{\mystoreparindent}{\the\parindent}
	\setlength{\parindent}{#1}
}{%
	\setlength{\parindent}{\mystoreparindent}
}

\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}

@@

\begin{document}
\bibliographystyle{apacite}
%\sffamily
%\maketitle

%Word count: 9263

\begin{singlespace}

\begin{myparindent}{0pt}
Short title: Phonetic transfer\\

Full title: Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals\\

Authors:\\

Auromita Mitra

EFL University,

Hyderabad,

India\\

Indranil Dutta

Jadavpur University,

Kolkata,

India\\

Acknowledgments:\\
We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.\\

Address for correspondence:\\
Auromita Mitra,

EFL University,

Hyderabad- 500007,

auromita.mitra@gmail.com

\end{myparindent}

\end{singlespace}

\newpage
\begin{myparindent}{0pt}
Abstract:\\
This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.  We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs. mixed productions. Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}. We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time. We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study. Results reveal no difference across paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\

Keywords: phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali

\end{myparindent}

\newpage

\section{Introduction: Phonetic transfer}

Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}. However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed. Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
%show cross-language-- rephrase

%keep just the types here, and put the next 2 lines in the causes&effects section?
Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production}; 
(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}. %These compare productions during mixed-language use to participants' own productions while using a single language. The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness. 
The present study is concerned with the latter. In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use these interchangeably here to indicate any interaction between two sets of phonetic norms.

Existing research on the phonetic effects of mixed-language production has largely focused on a limited set of phonologically related languages. A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer. However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants. This emphasizes the importance of considering data from a wider variety of
%elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer. 'Types of features'--separate line?- Brief mention - discursive factors in VOT. 
populations, language pairs,  and phonetic features in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
%line--rephrase
behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing. However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.

The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems. \ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs. the local act of switching between languages. Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter. The design of the present study is motivated on the basis of this discussion. A logical consequence of this model is that such phonetic interaction must be highly localized. \ref{duration} elaborates on this by reviewing literature on the duration and `reversibility' of transfer effects. A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies. In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that mediate such phonetic interaction (or fail to do so). We argue that these results point to an urgent need for expanding the range of language pairs and sound categories examined in order to make meaningful generalizations about the nature of short-term phonetic transfer. Following from this, \ref{asymmetry between sounds} discusses reported asymmetries between different sound categories of a single language, and develops the hypotheses of the present study on the basis of these. In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching. \ref{questions_and_hypotheses} summarizes the research questions and hypotheses.

@@

10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks. All the participants speak other languages in addition to Bengali and English. At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population. Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW (listening, speaking, reading, writing) skills in L1 and L2 across participants, and were compensated for their time.
